---
title: "Diagnosing failed R package installations"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Diagnosing failed R package installations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(sysreqr)
```

## When the preflight check did not happen

Sometimes preflight checks do not happen. A user runs `install.packages()`, a
package compiles from source, and the installation fails. The useful next step
is to turn the error into a small list of likely system packages, not to ask
the user to read a long compiler log alone.

`sysreqr` has two diagnosis paths:

* **Direct log patterns**, for common missing headers, linker errors, and
  `pkg-config` failures.
* **Failed R package lookup**, where the failed R package names are matched
  against the bundled or optional live system requirement database.

Both paths return a regular `sysreqr_plan`, so the output can be passed to
the same command, report, script, Dockerfile, GitHub Actions, and
administrator helpers used for preflight checks.

## Check the last error

After a failed install in the current R session:

```{r, eval = FALSE}
install.packages("xml2")
check_error(platform = "ubuntu-22.04")
```

`check_error()` reads `geterrmessage()` by default. You can also pass text
explicitly:

```{r}
check_error(
  text = "ERROR: configuration failed for package 'xml2'",
  platform = "ubuntu-22.04",
  backend = "bundled"
)
```

## Diagnose a log file

Use `diagnose_log()` for an install log on disk:

```{r, eval = FALSE}
plan <- diagnose_log("install.log", platform = "ubuntu-22.04")
```

Use the `text` argument when the log text is already in R:

```{r}
diagnose_log(
  text = "fatal error: libxml/parser.h: No such file or directory",
  platform = "ubuntu-22.04"
)
```

`diagnose_install_log()` is an alias for `diagnose_log()`.

## Start from failed package names

If the user already knows which R packages failed, skip log parsing:

```{r}
diagnose_failed_packages(
  c("xml2", "curl"),
  platform = "ubuntu-22.04",
  backend = "bundled"
)
```

This is useful when the log only says that an R package had non-zero exit
status. It is a medium-confidence result, because package installation can
fail for reasons unrelated to system libraries.

## Turn diagnosis into action

Diagnosis returns a `sysreqr_plan`. Use the same helpers as in the preflight
workflow:

```{r, eval = FALSE}
plan <- diagnose_log("install.log", platform = "ubuntu-22.04")
install_command(plan)
write_install_script(plan, "install-sysreqs.sh")
admin_request(plan)
write_report(plan, "SYSREQS.md")
```

For automation:

```{r, eval = FALSE}
dockerfile(plan)
github_actions(plan)
write_json(plan, "sysreqs.json")
```

For explanation:

```{r, eval = FALSE}
explain(plan)
as_data_frame(plan)
```

## Check a whole project

When one package fails, dependencies nearby in the project may fail next.
`check_project()` scans a project before the next installation attempt. It
reads `renv.lock`, then `DESCRIPTION`, then source files.

```{r, eval = FALSE}
project_plan <- check_project(".")
```

Include `Suggests` when preparing a development or testing environment:

```{r, eval = FALSE}
project_plan <- check_project(".", include_suggests = TRUE)
```

To inspect only package detection:

```{r, eval = FALSE}
detect_project_packages(".")
detect_project_packages(".", include_suggests = TRUE)
```

## Check an installed library

`check_library()` checks installed packages in an R library. This is useful
when a machine already has many packages installed and you want to see whether
their common external requirements are present.

```{r, eval = FALSE}
check_library()
check_library(c("xml2", "curl"))
```

## Interpret confidence

`sysreqr` uses conservative confidence labels:

* `high`: a direct missing-header or linker pattern, or structured upstream
  requirement data.
* `medium`: a likely tool failure, or a failed package lookup against the
  bundled fallback.
* `low`: reserved for future broader heuristics.

A medium-confidence result is still useful, but it is not proof that a missing
system library caused the failure. R version mismatches, network problems,
locked libraries, compiler bugs, unsupported package versions, and permission
problems can all produce failed installs that are outside the system
requirement database.

## When setup advice is the better answer

If repeated installs fail, run `setup_advice()` instead of chasing one error
at a time:

```{r, eval = FALSE}
setup_advice(
  packages = c("xml2", "curl"),
  platform = "ubuntu-22.04",
  script = "setup-sysreqr.sh"
)
```

That gives the user a broader checklist: binary R packages, source build
tools, optional R Project operating system repositories, and package-specific
system requirements.

## See also

* `vignette("preflight-setup")` for the preventive workflow.
* `vignette("faq")` for common diagnosis questions.
* `vignette("linux-fundamentals")` for the underlying GNU/Linux concepts.
