---
title: "Preflight setup before installing R packages"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Preflight setup before installing R packages}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(sysreqr)
```

## Why preflight?

A new GNU/Linux user often meets system requirements only after something has
already failed. The better workflow is to check the setup first, use binary R
packages where possible, and install source build tools only when source
packages are needed.

`sysreqr` is a zero-runtime-dependency helper for that workflow. It does not
run `sudo`, edit operating system repositories, or install system packages. It
prints commands and writes files that the user, or their administrator, can
review.

## Understand the platform

`sysreqr` tries to detect the current platform automatically, but examples
and reproducible scripts should usually pass a platform string.

```{r}
detect_package_manager("ubuntu-24.04")
resolve_platform("noble")
```

Common platform strings:

| Distribution | `<distro>-<version>`        | Codename alias |
|--------------|-----------------------------|----------------|
| Ubuntu       | `ubuntu-22.04`              | `jammy`        |
| Ubuntu       | `ubuntu-24.04`              | `noble`        |
| Ubuntu       | `ubuntu-26.04`              | `resolute`     |
| Debian       | `debian-12`                 | `bookworm`     |
| Debian       | `debian-13`                 | `trixie`       |
| RHEL family  | `rockylinux-9`, `redhat-9`  |                |
| Fedora       | `fedora-40`                 |                |
| openSUSE/SLE | `opensuse156`, `sle-15.6`   |                |

## Start with setup advice

`setup_advice()` produces a practical checklist for a Linux platform.

```{r, eval = FALSE}
setup_advice(platform = "ubuntu-24.04")
```

The advice has four layers:

1. **Posit Package Manager**, when a binary repository URL is known.
2. **Source build tools** such as `r-base-dev`, compilers, `make`, and
   `pkg-config`.
3. **Optional R Project operating system repository commands** for supported
   Ubuntu, Debian, and Fedora systems.
4. **Package-specific system requirements** when package names are supplied.

The R Project operating system repository layer is optional because it changes
the operating system repository configuration. It is most relevant when the
system R version is too old. It is not the first fix for every missing library.

To include package-specific requirements:

```{r, eval = FALSE}
setup_advice(
  packages = c("xml2", "curl"),
  platform = "ubuntu-24.04"
)
```

To write a reviewable shell script:

```{r, eval = FALSE}
setup_advice(
  packages = c("xml2", "curl"),
  platform = "ubuntu-24.04",
  script = file.path(tempdir(), "setup-sysreqr.sh")
)
```

The script contains source build tools and package system requirements as
active commands. Optional operating system repository commands are included as
comments because they should be reviewed first.

## Use Posit Package Manager

Binary R packages are the simplest way to avoid source compilation problems.

Build a Posit Package Manager repository URL:

```{r}
ppm_repo(platform = "ubuntu-24.04")
```

Preview the `.Rprofile` lines that would point R at that repository:

```{r}
use_ppm(platform = "ubuntu-24.04", dry_run = TRUE)
```

`use_ppm()` does not edit files unless `dry_run = FALSE` and an explicit
`path` is supplied.

Posit Package Manager binaries do not install the runtime system libraries
a package needs, so check the requirements anyway. On some distributions a
community binary repository handles system dependencies automatically
instead: r2u for Ubuntu, cran2copr for Fedora, CRAN2OBS for openSUSE. See
`vignette("faq")` for links.

When network access is available, these helpers query Posit Package Manager
support and system requirement data live:

```{r, eval = FALSE}
check_ppm("ubuntu-22.04")
ppm_platforms()
ppm_sysreqs(c("xml2", "curl"), platform = "ubuntu-22.04")
```

## Check packages before installing

`check_packages()` returns a `sysreqr_plan`. The plan is a data frame with the
R package, system requirement, installable system package, install script,
platform, package manager, installed status when available, source,
confidence, and notes.

```{r}
plan <- check_packages(
  c("xml2", "curl"),
  platform = "ubuntu-22.04"
)
plan
```

Turn the plan into common outputs:

```{r}
install_command(plan)
```

```{r, eval = FALSE}
write_install_script(plan, file.path(tempdir(), "install-sysreqs.sh"))
dockerfile(plan)
github_actions(plan)
```

Write files for review or automation:

```{r, eval = FALSE}
write_report(plan, file.path(tempdir(), "SYSREQS.md"))
write_json(plan, file.path(tempdir(), "sysreqs.json"))
```

Create a message for an administrator:

```{r, eval = FALSE}
admin_request(plan)
```

Inspect or explain a plan:

```{r, eval = FALSE}
as_data_frame(plan)
as_install_plan(plan)
is_sysreqr_plan(plan)
explain(plan)
```

## Pick a backend

`check_packages()` accepts four backend modes.

* `backend = "auto"` uses bundled data for simple, known CRAN packages on
  `apt`, `dnf`/`yum`, `zypper`, and `apk` platforms, then Package Manager,
  then `pak` when possible.
* `backend = "bundled"` uses only the static database shipped with the
  installed `sysreqr` release. It carries package names for `apt`,
  `dnf`/`yum`, `zypper`, and `apk`.
* `backend = "ppm"` uses the Posit Package Manager API when network access is
  available.
* `backend = "pak"` uses `pak::pkg_sysreqs()` when `pak` is installed.

`sysreqr` has zero runtime dependencies. Optional live backends (`ppm`, `pak`)
are used only when requested and available.

## See also

* `vignette("diagnosing-failures")` for post-failure log diagnosis.
* `vignette("linux-fundamentals")` for a GNU/Linux primer.
* `vignette("docker-and-ci")` for container and CI workflows.
* `vignette("faq")` for frequently asked questions.
